Bayesian inference of prehistoric population dynamics from multiple proxies: a case study from north of the Swiss Alps

### 2022-12-09 manuscript/paper_revision_1.Rmd
Corrections and comments
Comments in <!-- HTML comments -->. NB.: I haven't re-knitted anything.

title: "Bayesian inference of prehistoric population dynamics from multiple proxies: a case study from the North of the Swiss Alps"

@@

  Here, we present a Bayesian hierarchical model suite that uses Negative Binomial linear model and state-space representation to produce absolute estimates of past population size and density. At its core, the statistical model is as follows: if the proxies as a whole have a positive delta, a Negative Binomial draw for the new number of settlements is biased by a positive amount (and similarly for negative deltas). Using the  area North of the main ridge of the Swiss Alps in prehistoric times (6000–1000 BCE) as a case study, we show that combining multiple proxies (site counts, radiocarbon dates, dendrochronological dates, and landscape openness) produces a more robust reconstruction of population dynamics than any single proxy alone. The model's estimates of the credibility of its prediction, and the relative weight it affords to individual proxies through time, give further insights into the relative reliability of the evidence currently available for paleodemographic research. Our prediction of population development of the case study area accords well with the current understanding in the wider literature, but provides a more precise and higher-resolution estimate that is less sensitive to spurious fluctuations in the proxy data than existing approaches, especially the popular summed probability distribution of radiocarbon dates.

@@

  - We provide absolute estimates of population size and density on the area north of the Swiss Alpes in prehistoric times (6000–1000 BCE) based on an initial value

@@

@@

Prehistoric demography has experienced a resurgence in interest in recent years [@shennan_population_2000; @riede_climate_2009 and others in same issue], partly explained by a renewed interest in human–environment relations and human impact, necessarily requiring an assessment of population size. Kintigh et al. [-@kintigh_grand_2014] list human influence, dominance, population size, and population growth amongst their 'grand challenges' for archaeology in the 21st century.

In particular, the 'dates as data' technique [@rick_dates_1987], using the frequency of radiocarbon dates as a proxy for population dynamics, has been significantly developed in the last decade [e.g. @shennan_regional_2013] and widely applied to archaeological contexts worldwide [@crema_review_2022]. This approach has contributed greatly to our understanding of prehistoric demography, but is not without its critics [@attenbrow_hiscock_2015; @carleton_groucutt_2020; @price_end__end_2021]. While the methodology continues to evolve and address these critiques [@crema_review_2022], it remains subject to fundamental problems. We believe that these problems cannot be overcome by methodological refinements in this area alone. One of those problems is, that most analysis of this kind rely on a single proxy [@french_manifesto_2021; @schmidt_approaching_2021]. Other data for demographic processes, such as age composition of grave groups [@bocquet_appel_explaining_2008; @kohler_long_2014] exist, as does the use of environmental data [@lechterbeck2014]. The combination of a large number of indicators can compensate for the weaknesses of the individual proxies used and open up completely new possibilities for insights [@wolpert_past_2021, 10]. A Bayesian approach offers a robust, quantitative methodology for inferring prehistoric population dynamics from multiple proxies, including summed radiocarbon dates. Problems such as statistical dispersion of sum calibration and effects of the calibration curve will be mitigated respectively detected by this combination with indicators that are not subject to these effects. Moreover, in our approach, the mutual evaluation of the proxies in terms of concordance penalises divergent behaviour of an indicator (due to intrinsic biases) by reducing its relevance for the resulting estimate. While this does not eliminate these problems entirely, it does reduce them in a numerically manageable way for large data sets.

In the following article we present our methodology and demonstrate its application with a case study in the area north of the Swiss Alps. We use a type of Bayesian Generalised Linear Model [@kruschke2015] to combine the different proxies. Their assessment (weight and credibility intervals) results from the modelling and the combination of the data itself and is not defined a priori. Furthermore, we pursue the concept of a state space model [@auger-methe_guide_2021], a time series model in which a time series is interpreted as the result of a noisy observation of a stochastic process. We evaluate the number of settlements at time t as a function of the number of settlements at time t-1, modified according to the changes in the proxies. For this purpose, we also use these as the difference between these points in time and transform them accordingly. In this sense, it is not a deductive model for testing a hypothesis, but a reconstructive (abductive) one, such it is also used in environmental reconstructions [@inkpen_explaining_2009]. In this respect, we hope to convince the readers that such an approach is far better suited to realising reliable reconstructions of past population developments than any of the previous approaches.

@@

Our case study targets the period between 6000–1000 BCE. The lower limit of this time window was chosen to avoid the so-called 'Hallstatt plateau' in the Northern Hemisphere radiocarbon calibration curve, which causes difficulties for the using sum calibration and similar approaches. The upper limit coincides with post-glacial changes in pollen spectra, before which the openness indicator is highly unlikely to reflect human influence.

@@

The dataset we use for the number of dendro-dated wetland settlements in the Three Lakes region was collected by Julian Laabs for his PhD thesis [@laabs2019]. The time series used here runs from `r abs(min(dendro$date))` to `r abs(max(dendro$date))` BCE, and contains the number of chronologically registered fell phases at individual settlements. This results in a time series that reflects the settlement of the lakeshores in the Neolithic and Bronze Age periods.

## Radiocarbon Based Settlement Count

Summing up the probabilities of radiocarbon dates can undoubtedly be considered the standard approach in demographic reconstruction in archaeology today. In our variant of the summation calibration, we sum up the individual data in each site so that each site receives a weight of 1. These individual probabilities are then summed up over all sites, resulting in a radiocarbon based settlement count (RBSC). In many respects, this follows the same logic as generally applied summation calibration, where binning achieves much the same outcome.

@@

The dataset for the RBSC primarily consists of data from the XRONOS database (https://xronos.ch), supplemented by dates from the unpublished PhD thesis of Julian Laabs [@laabs2019] and the data collection of Martínez-Grau et al. [-@martínez-grau2021]. It contains a total of `r nrow(c14)` single ^14^C data from `r length(unique(c14$site))` sites (see Figure \@ref(fig:c14map)). The dates in the dataset range in ^14^C years from `r max(c14$bp)` to `r min(c14$bp)` uncal BP. This time window extends beyond the study horizon in order to minimise boundary effects.

@@

Dendrodated lakeshore settlements and RBSC are two fairly direct proxies for settlement numbers over time. However, both are subject to certain constraints, especially on the Swiss Plateau, but also beyond. They require the discovery of corresponding settlement sites. This is subject to preservation and discovery filters, which distort the results of the derived estimates. These biases are not connected to past population trends, but result from research activity, preservation and publication status, as well as dating strategies. Therefore, in our model we supplement these two direct proxies with such proxies that are more independent of these biases.

@@

To add another archaeological indicator of occupation, we include relative dating information obtained from the Swiss cantonal archaeology/heritage management authorities (Figure \@ref(fig:aoristmap)). These are primarily derived from scattered surface finds, which often have a low dating accuracy. We incorporate this data into our model as a typologically dated aorist time series. The dating accuracy is only in the range of archaeological periods, but the advantage is that we are not bound to the conditions and problems of radiocarbon dating and thus methodological issues of sum calibration can be avoided (scatter, calibration curve effects). Furthermore, this data provides an independent indicator with regard to the methodology of the ^14^C data, even if they are influenced by similar transmission filters and archaeological conditions as the evaluation of ^14^C data. Data from `r nrow(aorist)` sites were included in the aoristic sum, which is a very rough indicator due to the low dating accuracy offered by archaeological phases, but which nevertheless has an important role in the normalisation of the data due to its independence from calibration effects.

@@

@@

RBSC, openness and the dendro-dated settlement data was smoothed by a moving average with a 50 years window, corresponding to the unified sampling interval for all proxies. The choice of interval reflects a trade-off between the data quality of the proxies and the complexity of the model. The aoristic sum was not smoothed, because it already has a very coarse temporal resolution. In the construction of our 'observational model', we considered all these proxies as informative of the number of settlements located in the north of the Swiss Alps. Population development is simulated in a 'process model' using a Negative Binomial process.

@@

A Poisson distribution has the constraint that the variance is equal to the mean. This distribution may have been empirically very useful for modelling respective processes. However, we were advised in the peer review of the article that this limitation may not be justified. Therefore, we have rebuilt and run the model using a Negative Binomial Distribution, and present here the results of this process model, contrasting them with those of the Poisson model.

@@

Data collection were conducted as part of the project 'Beyond lake settlements' in the doctoral thesis of Julian Laabs, funded by the SNF (project number 152862, PI Albert Hafner) and as part of the XRONOS project, also funded by the SNSF (project number 198153, PI Martin Hinz). The development of the openness index took place within the framework of the project Time and Temporality in Archaeology (project number 194326, PI Caroline Heitz), inspired by the cooperation within the project QuantHum (project 169371, PI Marco Conedera), both also funded by the SNSF. Jan Kolář was supported by a long-term research development project (RV67985939) and by a grant from the Czech Science Foundation (19-20970Y). Julian Laabs was also funded by the German Research Foundation (DFG), Collaborative Research Centre 1266 'Scales of Transformation' (project number 2901391021). We also thank the Institute of Archaeological Sciences of the University of Bern for its support and faith in the outcome of our modelling project. Finally, we thank Michael Holton Price and W. Christopher Carleton for their really thorough review, helpful comments and valuable suggestions, which will certainly improve this manuscript significantly.

>>>>

title: "Bayesian inference of prehistoric population dynamics from multiple proxies: a case study from north of the Swiss Alps"

@@

  <!-- JR: For me this is far too much methodological detail for an abstract; I prefer the submitted version. I think simply removing "regression" is sufficient to head-off potential for confusion with an explanatory model. -->Here, we present a Bayesian hierarchical model suite that uses negative binomial linear model and state-space representation to produce absolute estimates of past population size and density. At its core, the statistical model is as follows: if the proxies as a whole have a positive delta, a negative binomial draw for the new number of settlements is biased by a positive amount (and similarly for negative deltas). Using an area north of the main ridge of the Swiss Alps in prehistoric times (6000–1000 BCE) as a case study, we show that combining multiple proxies (site counts, radiocarbon dates, dendrochronological dates, and landscape openness) produces a more robust reconstruction of population dynamics than any single proxy alone. The model's estimates of the credibility of its prediction, and the relative weight it affords to individual proxies through time, give further insights into the relative reliability of the evidence currently available for paleodemographic research. Our prediction of population development of the case study area accords well with the current understanding in the wider literature, but provides a more precise and higher-resolution estimate that is less sensitive to spurious fluctuations in the proxy data than existing approaches, especially the popular summed probability distribution of radiocarbon dates.

@@

  - We provide absolute estimates of population size and density on an area north of the Swiss Alps in prehistoric times (6000–1000 BCE) based on an initial value

@@

<!-- JR:  Okay I know I'm reopening an old issue here, but I really don't think "the north of the Swiss Alps" works as a geographic descriptor in English. If I encountered it for the first time I'd think it meant all of Europe north of the Alps. Removing the definite article helps a bit, but it's still confusing and awkward. I preferred "Swiss plateau" but, if that's unacceptable, can we just say "Northeast Switzerland" or something? -->

@@

Prehistoric demography has experienced a resurgence in interest in recent years [@shennan_population_2000; @riede_climate_2009 and others in same issue], partly explained by a renewed interest in human–environment relations and human impact, necessarily requiring an assessment of population size. Kintigh et al. [-@kintigh_grand_2014] list human influence, dominance, population size, and population growth amongst their 'grand challenges' for archaeology in the 21st century. In particular, the 'dates as data' technique [@rick_dates_1987], using the frequency of radiocarbon dates as a proxy for population dynamics, has been significantly developed in the last decade [e.g. @shennan_regional_2013] and widely applied to archaeological contexts worldwide [@crema_review_2022]. This approach has contributed greatly to our understanding of prehistoric demography, but is not without its critics [@attenbrow_hiscock_2015; @carleton_groucutt_2020; @price_end__end_2021]. While the methodology continues to evolve to address these critiques, methodological refinements alone cannot overcome the fundamental issues that limit its use as a direct proxy for past population [@crema_review_2022]. Similarly, while there are several other archaeological [e.g. @bocquet_appel_explaining_2008; @kohler_long_2014] and environmental [@lechterbeck2014] proxies for demographic processes, analyses that rely on any one of these proxies alone are inevitably limited by the nature of that proxy [@french_manifesto_2021; @schmidt_approaching_2021].

<!-- JR: I've replaced "indicator" with "proxy" in this section, but I think it's still mixed elsewhere. I don't mind which one we use, but I think we should stick to one term. -->

In this article we present a new, Bayesian methodology that combines multiple proxies to infer prehistoric population dynamics. As a case study, we apply this method to a region north of the Swiss Alps, deriving an absolute estimate of regional population density between 6000 and 1000 BCE from a combination of radiocarbon, dendrochronology, aoristic and landscape openness data. This Bayesian approach offers a robust, quantitative methodology for demographic inference that not only compensates for the weaknesses of individual types of record, but can open up entirely new areas of insight [@wolpert_past_2021, 10]. Issues inherent to single proxies (e.g. for radiocarbon, statistical dispersion from summing and distortions transferred from the calibration curve) are detected by cross-comparison with proxies that are not subject to these effects. The Bayesian algorithm penalises divergent behaviour in any one proxy (due to intrinsic biases) by reducing its relevance for the resulting estimate. While this does not eliminate these problems entirely, it does mitigate them in a numerically manageable way for large data sets.
<!-- JR: Not convinced that the following belongs in the introduction. Don't we say all this in the methods section? --->
Specifically, our method uses a type of Bayesian generalised linear model [@kruschke2015] to combine different proxies of past population density. Their assessment (weight and credibility intervals) results from the modelling and the combination of the data itself and is not defined a priori. Furthermore, we pursue the concept of a state space model [@auger-methe_guide_2021], a time series model in which a time series is interpreted as the result of a noisy observation of a stochastic process. We evaluate the number of settlements at time t as a function of the number of settlements at time t-1, modified according to the changes in the proxies. For this purpose, we also use these as the difference between these points in time and transform them accordingly. In this sense, it is not a deductive model for testing a hypothesis, but a reconstructive (abductive) one, such it is also used in environmental reconstructions [@inkpen_explaining_2009]. 
<!-- JR: end -->
In this respect, we hope to convince the readers that such an approach is far better suited to realising reliable reconstructions of past population developments than any of the previous approaches.

@@

Our case study targets the period between 6000–1000 BCE. The lower limit of this time window was chosen to avoid the so-called 'Hallstatt plateau' in the northern hemisphere radiocarbon calibration curve, which causes difficulties for the using sum calibration and similar approaches. The upper limit coincides with post-glacial changes in pollen spectra, before which the openness indicator is highly unlikely to reflect human influence.

@@

We used a dataset containing the number of dendro-dated wetland settlements in the Three Lakes region [@laabs2019].<!-- JR: As Julian is a co-author, I don't think it's necessary to attribute him specially in the text? --> The time series used here runs from `r abs(min(dendro$date))` to `r abs(max(dendro$date))` BCE, and contains the number of chronologically registered fell phases at individual settlements. This results in a time series that reflects the settlement of the lakeshores in the Neolithic and Bronze Age periods.

## Radiocarbon-based settlement count

We obtained radiocarbon dates for the study region from the XRONOS database (https://xronos.ch), supplemented by dates collected by Laabs [-@laabs2019] and Martínez-Grau et al. [-@martínez-grau2021]. The dataset contains a total of `r nrow(c14)` single ^14^C data from `r length(unique(c14$site))` sites (see Figure \@ref(fig:c14map)). The dates range from `r max(c14$bp)` to `r min(c14$bp)` uncal BP – corresponding to a calibrated range wider than our target of 6000–1000 BCE in order to minimise boundary effects.

<!-- JR: I'm not convinced about the remainder of this section or by the phrase "radiocarbon-based site settlement". Basically we are following the standard summed radiocarbon method, and calling it something different here is likely to confuse many. And while I agree with the intepretation in terms of probability theory, it seems out of place here. -->
Following the standard approach to sum calibration, we summed up the individual data in each site so that each site receives a weight of 1. These individual probabilities are then summed up over all sites, resulting in a radiocarbon based settlement count (RBSC). In many respects, this follows the same logic as generally applied summation calibration, where binning achieves much the same outcome.

@@

@@

<!-- JR: I am not sure the following paragraph is necessary, considering we have already made the case for "single proxy bad, multi-proxy good" several times. If we do keep it, it would perhaps fit better at the top of materials section (i.e., explaining the choice of proxies before we explain what they are). -->
Dendrodated lakeshore settlements and RBSC are two fairly direct proxies for settlement numbers over time. However, both are subject to certain constraints, especially on the Swiss Plateau, but also beyond. They require the discovery of corresponding settlement sites. This is subject to preservation and discovery filters, which distort the results of the derived estimates. These biases are not connected to past population trends, but result from research activity, preservation and publication status, as well as dating strategies. Therefore, in our model we supplement these two direct proxies with such proxies that are more independent of these biases.

@@

We obtained relative dating information from the Swiss cantonal archaeology/heritage management authorities (Figure \@ref(fig:aoristmap)). This is primarily derived from scattered surface finds, which often have a low dating accuracy. We incorporate this data into our model as a typologically dated aorist time series. The dating accuracy is only in the range of archaeological periods, but the advantage is that we are not bound to the conditions and problems of radiocarbon dating and thus methodological issues of sum calibration can be avoided (scatter, calibration curve effects). Furthermore, this data provides an independent indicator with regard to the methodology of the ^14^C data, even if they are influenced by similar transmission filters and archaeological conditions as the evaluation of ^14^C data. Data from `r nrow(aorist)` sites were included in the aoristic sum, which is a very rough indicator due to the low dating accuracy offered by archaeological phases, but which nevertheless has an important role in the normalisation of the data due to its independence from calibration effects.<!-- JR: Okay, really nitpicking here, but... are they independent of calibration effects? We generally assign absolute dates to periods by interpreting radiocarbon chronologies and if, say, the boundary between two periods falls on a calibration plateau, the date ranges for those periods are going to be correspondingly imprecise, no? --->

@@

<!-- JR: So I feel like we've ended up with a lot more *technical* detail in the preceeding four sections, when what the reviewers asked for was a more thorough justification of *why* they're population proxies. I don't think this is necessarily an improvement... -->

@@

RBSC, openness and the dendro-dated settlement data was smoothed by a moving average with a 50 years window, corresponding to the unified sampling interval for all proxies. The choice of interval reflects a trade-off between the data quality of the proxies and the complexity of the model. The aoristic sum was not smoothed, because it already has a very coarse temporal resolution. In the construction of our 'observational model', we considered all these proxies as informative of the number of settlements located north of the Swiss Alps. Population development is simulated in a 'process model' using a Negative Binomial process.

@@

A Poisson distribution has the constraint that the variance is equal to the mean. This distribution may have been empirically very useful for modelling respective processes. However, we were advised in the peer review of the article that this limitation may not be justified. Therefore, we have rebuilt and run the model using a negative binomial Distribution, and present here the results of this process model, contrasting them with those of the Poisson model.

@@

Data collection were conducted as part of the project 'Beyond lake settlements' in the doctoral thesis of Julian Laabs, funded by the SNF (project number 152862, PI Albert Hafner) and as part of the XRONOS project, also funded by the SNSF (project number 198153, PI Martin Hinz). The development of the openness index took place within the framework of the project Time and Temporality in Archaeology (project number 194326, PI Caroline Heitz), inspired by the cooperation within the project QuantHum (project 169371, PI Marco Conedera), both also funded by the SNSF. Jan Kolář was supported by a long-term research development project (RV67985939) and by a grant from the Czech Science Foundation (19-20970Y). Julian Laabs was also funded by the German Research Foundation (DFG), Collaborative Research Centre 1266 'Scales of Transformation' (project number 2901391021). We also thank the Institute of Archaeological Sciences of the University of Bern for its support and faith in the outcome of our modelling project. Finally, we would like to thank reviewers Michael Holton Price and W. Christopher Carleton, as well as editor Enrico Crema, for an extremely thorough review process. Their helpful comments and valuable suggestions have improved both this paper and the model it describes signficiantly.
